---
title: "CDS workflow functionality"
author: "Koen Hufkens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CDS workflow functionality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# load the library
library(ncdf4)
```

## Running workflows on Copernicus' Climate Data Store

[Copernicus.eu](https://www.copernicus.eu/en) provides a set of interesting data sets for research, education, and applied earth sciences on their [Climate Data Store](https://cds.climate.copernicus.eu) (CDS). However, in many cases through the conventional API you can only download the raw data at their native temporal resolution.

Many people however don't require hourly or 6-hourly data for their research purposes. Using the standard API therefore results in queries of large amount of data which then need to be temporally down-sampled to get, for example, daily mean temperature values.

Many of these issues can be side stepped by using the [CDS Toolbox workflows](<https://cds.climate.copernicus.eu/toolbox/doc/api.html>). These allow you to execute `python` code on CDS servers, as provided through an R script.

## The Workflow syntax

Below we'll use a CDS Toolbox python script to extract a time series for a variable at a given (point) location. This is a common operation for people working with point (site/location) data.

On the CDS website the example script `03 Extract time series and plot graph` in the toolbox editor allows you to extract a daily mean time series for a given (defined) variable and a given location. The code below is normally used within the context of the CDS Toolbox Editor platform. However, we can reuse the same code within the context of a `ecmwfr` workflow request.

The code below lists all available variables to consider in combination with an API call defining the data ranges available (both spatial and temporal). This example is limited to the years 2008 - 2017 and 6-hourly time steps (although more data is available).

We will make one minor change to the original script which is critical for the success of our workflow. Instead of returning a figure, we will want to return the data on which the figure is based. We therefore changed the last line of the original file to `return data_daily`. We can save this script as a normal python script in any editor. For this example the data was saved into a file called `era5.py`

``` python
import cdstoolbox as ct

layout = {
    'input_ncols': 3,
}

variables = {
    'Near-Surface Air Temperature': '2m_temperature',
    'Eastward Near-Surface Wind': '10m_u_component_of_wind',
    'Northward Near-Surface Wind': '10m_v_component_of_wind',
    'Sea Level Pressure': 'mean_sea_level_pressure',
    'Sea Surface Temperature': 'sea_surface_temperature',
}


@ct.application(title='Extract a time series and plot graph', layout=layout)
@ct.input.dropdown('var', label='Variable', values=variables.keys(), description='Sample variables')
@ct.input.text('lon', label='Longitude', type=float, default=75., description='Decimal degrees')
@ct.input.text('lat', label='Latitude', type=float, default=43., description='Decimal degrees')
@ct.output.livefigure()
def plot_time_series(var, lon, lat):
    """
    Application main steps:

    - set the application layout with 3 columns for the input and output at the bottom
    - retrieve a variable over a defined time range
    - select a location, defined by longitude and latitude coordinates
    - compute the daily average
    - show the result as a timeseries on an interactive chart

    """

    # Time range
    data = ct.catalogue.retrieve(
        'reanalysis-era5-single-levels',
        {
            'variable': variables[var],
            'grid': ['3', '3'],
            'product_type': 'reanalysis',
            'year': [
                '2008', '2009', '2010',
                '2011', '2012', '2013',
                '2014', '2015', '2016',
                '2017'
            ],
            'month': [
                '01', '02', '03', '04', '05', '06',
                '07', '08', '09', '10', '11', '12'
            ],
            'day': [
                '01', '02', '03', '04', '05', '06',
                '07', '08', '09', '10', '11', '12',
                '13', '14', '15', '16', '17', '18',
                '19', '20', '21', '22', '23', '24',
                '25', '26', '27', '28', '29', '30',
                '31'
            ],
            'time': ['00:00', '06:00', '12:00', '18:00'],
        }
    )

    # Location selection

    # Extract the closest point to selected lon/lat (no interpolation).
    # If wrong number is set for latitude, the closest available one is chosen:
    # e.g. if lat = 4000 -> lat = 90.
    # If wrong number is set for longitude, first a wrap in [-180, 180] is made,
    # then the closest one present is chosen:
    # e.g. if lon = 200 -> lon = -160.
    data_sel = ct.geo.extract_point(data, lon=lon, lat=lat)

    # Daily mean on selection
    data_daily = ct.climate.daily_mean(data_sel)

    fig = ct.chart.line(data_daily)

    #return fig
    return data_daily
```

With the script adjusted and saved we can now use it in our `ecmwfr` workflow run as follows.

First we read in the python script as a long string into R. Use `readlines()` and collapse the various lines using the appropriate `\n` separator.

``` r
code <- readLines("era5.py") |>
  paste0(collapse = "\n")
```

We then have to formulate a query which sets the right variables in the python script, mainly the parameters `var`, `lat` and `lon`. The request is a nested list with a `code` argument, containing the code you want to run, a `kwargs` variable containing a named list of variables to forward to the code, a `workflow_name` defining the python application you want to call (in this case `plot_time_series`, and a `target` output variable name setting the filename of the resulting output.

With a correct query specified we can now run this workflow request using a normal `wf_request()` call, which will run the code on the CDS server and return the result to the target file.

``` r
# A query for 2m surface temperature
request = list(
  code = code,
  kwargs = list(
      var = "2m_temperature",
      lat = 50,
      lon = 20
  ),
  workflow_name = "plot_time_series", # name of the python subroutine / app
  target = "test.nc"
)

# download the data
file <- wf_request(
    user = USER_ID,
    request = request,
    path = tempdir()
  )
```

Once the file is downloaded we can open this file using the `ncdf4` library. Depending on the python script used you will be returned either `netCDF` formatted geospatial data (i.e. maps) or non-geospatial data. In this case the returned data is not geospatial, a time-series for our desired location.

Below we open the `netCDF` file and read in the temperature (t2m) data into variable `t`.

``` r
# open the netcdf file and print the meta-data
f <- ncdf4::nc_open(file.path(tempdir(),"test.nc"))
print(f)

# read in the temperature data stored in field "t2m"
# and the dates from the "time" field
temp <- ncdf4::ncvar_get(nc = f, "t2m")
time <- ncdf4::ncvar_get(nc = f, "time")

# get the starting point of the time series
# and add the increments (time)
start <- ncdf4::ncatt_get(f, "time")$units
start <- as.Date(start, format = "days since %Y-%m-%d 00:00:00")
time <- start + time

# close the file once done
ncdf4::nc_close(f)
```

We can now plot this data

``` r
plot(time, temp, ylab = "Temperature (K)")
```

```{r echo = FALSE}
nc_file <- system.file(package = "ecmwfr","extdata/cds_workflow.nc")

# open the netcdf file and print the meta-data
f <- ncdf4::nc_open(nc_file)

# read in the temperature data stored in field "t2m"
# and the dates from the "time" field
temp <- ncdf4::ncvar_get(nc = f, "t2m")
time <- ncdf4::ncvar_get(nc = f, "time")

# get the starting point of the time series
# and add the increments (time)
start <- ncdf4::ncatt_get(f, "time")$units
start <- as.Date(start, format = "days since %Y-%m-%d 00:00:00")
time <- start + time

# close the file once done
ncdf4::nc_close(f)

# plot data
plot(time, temp, ylab = "Temperature (K)")
```
