---
title: "ecmwfr"
author: "Koen Hufkens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ecmwfr functionality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

opts <- options(keyring_warn_for_env_fallback = FALSE)

# load the library
library(ncdf4)
library(raster)
library(maps)
library(ecmwfr)

# import the encrypted key if desired
# (does not exist locally)
if(system("echo $KEY", intern = TRUE) != ""){
  wf_set_key(email = "khrdev@outlook.com",
             key = system("echo $KEY", intern = TRUE))
}

# check cran, same routine as skip_on_cran()
# but not dependent on testthat which might
# not be available on user systems (not required
# only suggested)
check_cran <- function() {
  if (identical(tolower(Sys.getenv("NOT_CRAN")), "true")) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}

# do cran check
cran <- check_cran()

# for local render set to true
# mainly important for rendering
# a website using pkgdown
#cran <- TRUE
```

# ecmwfr

Programmatic interface to the ['ECMWF' web API services](https://confluence.ecmwf.int/display/WEBAPI/ECMWF+Web+API+Home). Allows for easy downloads of ECMWF [public data](https://confluence.ecmwf.int/display/WEBAPI/Access+ECMWF+Public+Datasets).

## Setup

Before starting save your provided ECMWF key to your local keychain. In this example I use my R development account with the key hidden using Xs. The package does not allow you to use your key inline in scripts to limit security issues when sharing scripts on github or otherwise.

```{r eval = FALSE}
# set a key to the keychain
wf_set_key(email = "khrdev@outlook.com",
           key = "XXXXXXXXXXXXXXXXXXXXXX")

# you can retrieve the key using
wf_get_key(email = "khrdev@outlook.com")

# the output should be the key you provided
# in this case represented by the fake X string.
# "XXXXXXXXXXXXXXXXXXXXXX"
```

Before you can download any data you have to make sure to accept the terms and conditions here:
[https://apps.ecmwf.int/datasets/licences/general/](https://apps.ecmwf.int/datasets/licences/general/).

## Listing Ancillary data

You can list ancillary data such as user information, datasets and services using the following calls.

```{r eval = cran}
# user info
user_info <- wf_user_info(email = "khrdev@outlook.com")
head(user_info)

# services
services <- wf_services(email = "khrdev@outlook.com")
head(services)

# datasets
datasets <- wf_datasets(email = "khrdev@outlook.com")
head(datasets)
```

## Data Requests

To download data use the wf_request() function, together with your email and a request string syntax [as documented](https://confluence.ecmwf.int/display/WEBAPI/Brief+request+syntax#Briefrequestsyntax-Syntax). Instead of `json` formatting the function uses a simple `R` list for all the arguments.

```{r eval = cran}
# this is an example of a request
my_request <- list(stream = "oper",
                  levtype = "sfc",
                  param = "167.128",
                  dataset = "interim",
                  step = "0",
                  grid = "0.75/0.75",
                  time = "00",
                  date = "2014-07-01/to/2014-07-31",
                  type = "an",
                  class = "ei",
                  area = "73.5/-27/33/45",
                  format = "netcdf",
                  target = "tmp.nc")

# an example download using fw_request()
# using the above request list()
wf_request(
  email = "khrdev@outlook.com",
  transfer = TRUE,
  path = "~",
  request = my_request,
  verbose = FALSE)
```

This operation might take a while. A progress indicator will keep you informed on the status of your request. The download will expire after a default time out of one hour. You will be informed of where to track your job request on [the ECMWF website](https://apps.ecmwf.int/webmars/joblist/), and provided with a download url which can be sued with the `wt_transfer()` funtion to download the data on a later moment. Keep in mind that you can only have 3 active downloads and 20 job items queued! 

If the download is successful you can visualize one layer in the retrieved stack of temperatures at 2 meters using the plotting routine below.

```{r fig.width = 7, fig.height = 7, eval = cran}
s <- raster::stack("~/tmp.nc")
print(s)

raster::plot(s[[1]], main = "2 meter temperature (Kelvin)")
maps::map("world", add = TRUE)
```
